---
title: "Main Data to Global - `r date()`"
author: "Mike and Dave"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
    fig_caption: true
    fig_width: 9
    fig_height: 6
    toc_depth: 3
  html_document:
    code_folding: hide
    fig_width: 6
    fig_height: 6
    toc: truef
    fig_caption: true
  params:

---

# Documentation (README)

This file creates the raw .csv export and codebook of the main (i.e. non-Curacao) study data that we send to globa.l

Input:

* mdta_02w_v2024-09-24.rds
* "Variables to send to global...".xlsx with the fields to include  
* the "skeleton file" ../skel/char_child_par_fix_v16.xlsx, with the labels
* A google spreadsheet "Changes needed to LFF quant data to Global", with the manual changes to the data

Key Options:

* None  

Code Overview:

* Load the data from mdta...rds, extract desired variables from "Variables to send to global".xls
* Make any adjustments to the underlying data (only step in this program is to aggregate languages if fewer than 5 per site)
* Create an intitial codebook using the vt() function. If value labels are missing, read them in from the "skel" file. 
** This is particularly helpful for the checkbox fields
* Create a mirror data set with the factors as numeric. This is what we will export.
* Gather the value labels of the original factors. Run vt() on the numeric mirror data set to get values combine that information to get the data description for factor variables.
* Gather the field levels from the character variables.
* Import Manual fixes from Google sheets
* Clean up the code book and export it, and the data as .csv, .rds, .sav

Output:
* code book
* the data as .csv, .rds and .sav files


# Setup

```{r Setup, warning=FALSE}
# FILE PATHS AND PARAMETERS
rm(list=ls())
#lff1_raw_data_path="//ars-data-01.sde.net.ucsf.edu/MyResearchShared/LFranck_shared/RMHC/Data/mdta_02_v2024-09-24.rds"
lff1_raw_data_path="C:/ucsf/mdta_02_v2024-09-24.rds"
skeleton_file_w_variable_labels="char_child_par_fix_v16.xlsx"
#vars_to_send_to_global="Variables to send to global.xlsx"
vars_to_send_to_global="Variables to send to global - 18mar2025 10pm.xlsx"
google_sheet_of_manual_fixes="https://docs.google.com/spreadsheets/d/18Sl4ki7Sw7MB1rWnB1ZDFooTT99HAfyPKhHVHAyoJV0/edit?gid=1329229673#gid=1329229673"
sheet_tab_of_manual_fixes="Manual_Fixes"
codebook_output_path="LFF1_global_codebook.csv"
csv_data_output_path="LFF1_global_data.csv"

# LOAD PACKAGES. INSTALL IF NEEDED
if(!require("pacman")) {install.packages("pacman");library(pacman)}
pacman::p_load(tidyr,vtable,rio,readxl,writexl, sjlabelled,Hmisc,viscomplexr,lares,
               forstringr,googlesheets4,tibble, forcats)

```


## Load data

```{r data_prep}
# MIKE'S CODE TO CREATE DB FOR GLOBAL
# THANKS MIKE!

# data
mdta <- readRDS(lff1_raw_data_path)

# Variables to send to Global: only the s1 vars listed. have to check for s2 & s3
globalvars <- import(vars_to_send_to_global)
gvars <- globalvars$variable

## Checking for s2 & s3 variables
# separate out the variables without a round prefix
noprefix <- c("site","region2", "los")
include <- gvars[-which(gvars %in% noprefix)]

# just take the root of all the variables and then prefix the rounds
# DGS NOTE: depanx(3 times), housing3.factor (twice) and survlang (3 times)
# ALL APPEAR MORE THAN ONE.DO A UNIQUE
root <- substr(include, 4, nchar(include))
unique_root=unique(root)

# all possibilities of variable names to check for
allpos <- c(paste0("en_", unique_root), paste0("dc_", unique_root), paste0("pd_", unique_root))

# add back in the variables with no round prefix
allpos <- c(noprefix, allpos)
```

# Swap out if there is a factor version
Currently not evaluating this code. It's a good idea, as there are 58 variables with factor equivalents, but it ended up leading to the program crashing, and I couldn't find the tiem to figure out why.
```{r replace_if_factor_var_equivalent, eval=FALSE}
allpos_w_factor_equivalents=allpos
for(i in 1:length(allpos)){
  factor_name=paste0(allpos[i],".factor")
  allpos_w_factor_equivalents[i]=ifelse(factor_name %in% names(mdta),factor_name,allpos[i])
}
```

# Subset to vars
```{r use_these_vars}
# get all variables in allpos which are actual variables in mdta
#globdata <-mdta %>% dplyr::select(any_of(allpos_w_factor_equivalents))
globdata <-mdta %>% dplyr::select(any_of(allpos))
```

# Aggregate langs if 5 or fewer at a site
```{r aggregate_langs}
# SHOW LANGUAGE DISTRIBUTION, BEFORE TURNING LESS FREQ LANGS TO OTHER
before= globdata |> 
  dplyr::group_by(site,en_survlang) |>
  dplyr::tally() |>
  spread(en_survlang,n)
print(n=15,before)

for (var in c("en_survlang","dc_survlang","pd_survlang")) {
  for (house in unique(globdata$site)) {
    few_speakers = globdata |>
      dplyr::filter(site==house) |>
      dplyr::group_by(!!rlang::sym(var)) |>
      dplyr::tally() |>
      dplyr::filter(n<=6) |>
      dplyr::select(-n)

langs_to_code_as_other=as.character(few_speakers)

globdata[globdata$site==house & globdata[,var] %in% langs_to_code_as_other,var]="ot"
  }
}

after= globdata |> 
  dplyr::group_by(site,en_survlang) |>
  dplyr::tally() |>
  spread(en_survlang,n)
print(n=15,after)
```


# Initialize codebook
```{r initiate_codebook}
#INITIAL CODEBOOK, WITH SOME HOLES TO FILL
codebook=vt(globdata, factor.limit=0, out="return")

#IDENTIFY UNLABELED VARIABLES, USE LABELS FROM THE SKELETON SPREADSHEET
xtra_labels=read_excel(skeleton_file_w_variable_labels,sheet="Sheet 1")
these_vars_need_label = codebook |> 
  dplyr::filter(Label=="NULL") |> 
  dplyr::select(Name)

for(var in these_vars_need_label$Name){
  lab = xtra_labels |> dplyr::filter(variable==var) |> dplyr::pull(question)
  if(length(lab)>1) lab=lab[1]
  eval(parse(text=paste0("Hmisc::label(globdata$",var,") <- lab")))
}


# # POSSIBLE UPGRADE: JUST USE VARIABLE LABELS FROM SKEL, IF THEY EXIST. ADD HERE
# # THEN HAVE 255 THAT STILL NEED A LABEL
# # BUT THIS LEAD TO CRASHING LATE ON. WEIRD
# #for(var in codebook$Name){
# for(var in these_vars_need_label$Name){
#   lab = xtra_labels |> dplyr::filter(variable==var) |> dplyr::pull(question)
#   if(length(lab)>1){
#     lab=lab[1]
#     eval(parse(text=paste0("Hmisc::label(globdata$",var,") <- lab")))
#     }
# }

#ONLY THE FIRST CHECKBOXES ARE LABELED. ADD LABELS TO THE OTHERS
#IF NO LABEL, BUT NAME HAS SAME PREFIX AS PREVIOUS VAR, USE LABEL OF PREVIOUS VAR
codebook=vt(globdata, factor.limit=0, out="return")
for (i in 2:nrow(codebook)) {
  if(substr(codebook[i,"Name"],1,6)==substr(codebook[i-1,"Name"],1,6) & is.na(codebook[i,"Label"])) {
    add_this_label=codebook[i-1,"Label"]
    var=codebook[i,"Name"]
    #CHANGE IN globdata AND IN codebook, THE CODEBOOK
    eval(parse(text=paste0("Hmisc::label(globdata$",var,") <- add_this_label")))
    codebook[i,"Label"]=add_this_label
  }
}

#FOR CHECKBOXES WE HAVE HAVE THE QUESTION, BUT NEED TO ADD THE CORRESPONDING  
#ANSWER WHICH IS IN skeleton_file_w_variable_labels. FIRST RERUN CODEBOOK
codebook=vt(globdata, factor.limit=0, out="return")
for (i in 1:nrow(codebook)) {
  var=codebook[i,"Name"]
  if(grepl('___',var)) {
    answer = xtra_labels |> dplyr::filter(variable==var) |> dplyr::pull(level_label)
    add_this_label=paste0(codebook[i,"Label"]," ",answer)
    #CHANGE IN globdata AND IN codebook, THE CODEBOOK
    eval(parse(text=paste0("Hmisc::label(globdata$",var,") <- add_this_label")))
  }
}
codebook=vt(globdata, factor.limit=0, out="return")

# WE ARE STILL MISSING THE VARIABLE LABEL FOR dc_hrsathosp.r. FILL THIS IN
these_vars_still_need_label = codebook |> 
  dplyr::filter(is.na(Label)) |> 
  dplyr::select(Name)

for(var in these_vars_still_need_label$Name){
  lab = xtra_labels |> dplyr::filter(variable==var) |> dplyr::pull(variable_label)
  if(length(lab)>1) lab=lab[1]
  eval(parse(text=paste0("Hmisc::label(globdata$",var,") <- lab")))
}

#WE HAVE THE VARIABLE LABELS! MAYBE NOT THE VALUE LABELS
#FOR CLARITY, CREATE A COLUMN W/ THE DATA TYPE
codebook=vt(globdata, factor.limit=0, out="return")
codebook$Type=""
for (i in 1:nrow(codebook)){
  var=codebook[i,"Name"]
  var_type=class(globdata[,var])[2]
  codebook[i,"Type"]=var_type
}
```


# Code output data as numeric
```{r coding_factors}
globdata_w_factor_as_numeric=globdata

# TRUST ME IT WILL BE EASIER IF SITES DON'T HAVE CODES
codebook$Type [1]="Character"

vars_w_factor_in_name=codebook$Name[codebook$Type=="factor"]
for(var in vars_w_factor_in_name) {
  globdata_w_factor_as_numeric[,var]=as.numeric(globdata_w_factor_as_numeric[,var])
}

# CREATE A CODEBOOK W NUMERIC FACTORS, TO PULL THOSE NUMERIC VALUES, 
# THEN ADD THIS TO THE CODEBOOK
codebook_w_numeric=vt(globdata_w_factor_as_numeric, factor.limit=0, out="return")

codebook=codebook |>
  dplyr::select(-Class) |>
  dplyr::rename(Orig_Values=Values) |>
  dplyr::relocate(Type,.after = Name)

codebook=cbind(codebook,Values_After_Fac2Num=codebook_w_numeric$Values)

char_array_to_txt <- function(char_array) {
  concatenated_str_vec=vector2String(char_array)
  return(substr(concatenated_str_vec,3,nchar(concatenated_str_vec)-1))
}

#CREATE A FINAL COLUMN COMBINING THE NUMERIC FACTORS WITH THEIR LEVELS
codebook$Factor_levels=""
codebook$Data_Range=""
for (i in 1:nrow(codebook)){
  var=codebook[i,"Name"]
  if(codebook[i,"Type"]=="factor") {
  #if(grepl("factor",var)) {
    fac_levels=levels(globdata[,var])
    codebook[i,"Factor_levels"]=char_array_to_txt(fac_levels)
    codebook[i,"Data_Range"]=paste0(codebook[i,"Values_After_Fac2Num"],", corresponding to: ",codebook[i,"Factor_levels"])
  } else {
    codebook[i,"Data_Range"]=codebook[i,"Orig_Values"]
  }
}
```


# Character variables
```{r character_vars}
character_vars_wo_description=codebook |>
  dplyr::filter(Type=="character", Data_Range=="") |>
  dplyr::select(Name,Data_Range)

for(var in character_vars_wo_description$Name){
  var_cats=names(table(mdta[,var]))
  codebook[codebook$Name==var,"Data_Range"]= paste(var_cats,collapse = ", ")  
}
```

# Use value labels if there is a factor version
```{r check_if_factor_equivalents}
codebook$factor_equivalent
for(var in codebook$Name){
  factor_name=paste0(var,".factor")
  codebook[codebook$Name==var,"factor_equivalent"]=ifelse(factor_name %in% names(mdta),factor_name,NA)
}

codebook$factor_equivalent_levels=""
for (i in 1:nrow(codebook)){
  if(codebook[i,"Type"] %in% c("numeric","integer") & !is.na(codebook[i,"factor_equivalent"])){
    var_factor_equivalent=codebook[i,"factor_equivalent"]
    fac_levels=levels(mdta[,var_factor_equivalent])
    codebook[i,"factor_equivalent_levels"]=char_array_to_txt(fac_levels)
    codebook[i,"Data_Range"]=paste0(codebook[i,"Data_Range"],
                                          ", corresponding to: ", codebook[i,"factor_equivalent_levels"])
  }
}

```
# Manual fixes in Google Sheets
```{r google_sheet_manual_fixes, echo=FALSE, include=FALSE}
#Read google sheets data into R
manual_fixes <- read_sheet(google_sheet_of_manual_fixes,sheet=sheet_tab_of_manual_fixes)

for (i in 1:nrow(manual_fixes)) {
  var=manual_fixes[i,"Variable_Name"][[1]]
  if(!is.na(manual_fixes$New_Survey_Item[i])) {
    codebook[codebook$Name==var,"Label"]=manual_fixes[i,"New_Survey_Item"]
  }
    if(!is.na(manual_fixes$New_Data_Range[i])) {
    codebook[codebook$Name==var,"Data_Range"]=manual_fixes[i,"New_Data_Range"]
  }
}

```


# Clean up before exprot
```{r clean_up_before_export}

# THE ROW ORDER HAS GOTTEN SCREWY. RESTORE IT AND ONLY SAVE COLUMNS NEEEDED
codebook$row_num=row.names(codebook)

#JUST THE 4 COLUMNS. CLEAN UP THE NAMES? VARIABLE, TYPE, SURVEY ITEM, DATA INFORMATION
codebook=codebook |>
  dplyr::select(Name,Type,Label,Data_Range) |>
  dplyr::rename('Variable Name'=Name,'Data Type'=Type,'Survey Question/Item'=Label,
                'Variable Description'=Data_Range) 

# CHANGE LABELS FOR DATA TYPE
codebook$`Data Type`[codebook$`Data Type`=="Character"]="Character Field"
codebook$`Data Type`[codebook$`Data Type`=="character"]="Character Field"
codebook$`Data Type`[codebook$`Data Type`=="factor"]="Coded Category"
codebook$`Data Type`[codebook$`Data Type`=="numeric"]="Continuous Numeric"
codebook$`Data Type`[codebook$`Data Type`=="integer"]="Integer Numeric"
```

# Output data and codebook
```{r output}
write.csv(globdata_w_factor_as_numeric,file=csv_data_output_path, row.names=F, na="")
write.csv(codebook, file=codebook_output_path, row.names=F)
#BECAUSE WE HAVE VARIABLES WITH "." IN THE NAMES, CAN'T WRITE AS .dta
#haven::write_dta(globdata, path=paste0(substr(codebook_output_path,1,nchar(codebook_output_path)-3),"dta"))
saveRDS(globdata, file=paste0(substr(codebook_output_path,1,nchar(codebook_output_path)-3),"rds"))
haven::write_sav(globdata, path=paste0(substr(codebook_output_path,1,nchar(codebook_output_path)-3),"sav"))
```

# Investigate a couple of variables
```{r check_a_few_vars, eval=FALSE}
country= mdta |> 
  dplyr::group_by(en_countryc.factor,en_countryc) |>
  dplyr::tally() |>
  spread(en_countryc,n)
print(n=15,country)

needs= mdta |> 
  dplyr::mutate(
    en_unmetneeds=fct_recode(en_needs_unmet.factor,
                             "Unmet needs at EN" = "Checked",
                             "No unmet needs at EN" = "Unchecked"),
    pd_unmetneeds=fct_recode(pd_needs_unmet.factor,
                             "Unmet needs at PD" = "Checked",
                             "No unmet needs at PD" = "Unchecked")
  ) |>
  dplyr::group_by(en_unmetneeds,pd_unmetneeds) |>
  dplyr::tally() |>
  spread(en_unmetneeds,n)
gt_1=gt::gt(needs)
gt_1
gt_2=gt::rm_stubhead(data=gt_1)
gt_2

```


